General constraints on magic distillation protocols from statistical mechanics

### 2021-05-17 magic_paper.tex
Smoothing and re-ordering. Few things still to do.

\nick{re-write based on new relative majorisation approach}
The past few years have seen rapid progress towards the goal of a fault-tolerant quantum computer~\cite{Fowler_2012, Herrera_2010, Nickerson_2014, Nikahd_2017, chao_2018, lin_pieceable_2020, Lin_2020, Bourassa_2021}. However, many challenges remain and there is increasing need for theory to take into account physical limitations of the hardware involved. The surface code~\cite{Bravyi_1998, Freedman_2001, Dennis_2002, Raussendorf_2007} is a leading framework for fault-tolerance with very high error thresholds. Within this scheme, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, we also know that it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally~\cite{Calderbank_1996, Steane_1996}, one needs to find ways around the Eastin-Knill restriction. This can be achieved by injecting in quantum states, called magic states, which promote the Clifford group to universal quantum computing~\cite{cit:bravyi}. The obstacle to this is that these states are invariably noisy and so protocols involving stabilizer operations must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question then arises about the overhead on purifying many copies of a magic state into less noisy forms.

@@

We approach this problem using tools from statistical mechanics and recent work from the overlap of thermodynamics and quantum information theory~\cite{cit:gour, cit:gour2}\nick{You had a specific review in mind I think, which one?}. In particular, we shall view \emph{magic as a form of free energy}, relative to the set of stabilizer states. This perspective takes the convex set of stabilizer states as the set of ``equilibrium'' states, and magic corresponds to a form of non-equilibrium free energy that is strictly non-classical.

@@

\begin{figure}[t]
    \centering
        \includegraphics[scale=0.4]{figs/protocol_diagram.pdf}
    \caption{\textbf{Action of a magic protocol} 
	A magic protocol aims to turn $n$ copies of some magic state $\rho_1$ to $m$ copies of some other state $\rho_2$.
	Constraints on the efficiency of the protocol can be imposed by considering its action on an separate probe state $\tau$.
	\nick{How do we improve the illustration?}
    }
    \label{fig:sketch}
\end{figure}

@@

We now consider a magic distillation protocol on multiple identical qudits in a noisy magic state $\rho(\epsilon)$, with noise parameter $\epsilon$, sending 
\begin{equation}
\rho(\epsilon)^{\otimes n} \longrightarrow \E(\rho(\epsilon)^{\otimes n}) =\rho(\epsilon')^{\otimes m}
\end{equation}
with $n \gg 1$ \nick{Should we replace $m$ with $R n$ or $n'$ throughout?}, $\epsilon' <\epsilon$ and $\E$ denoting the quantum channel induced by the protocol. We can now provide distillation bounds that depend on energetic and entropic aspects of the protocol. We do this relative to a reference temperature that can be chosen freely. The key idea is that the physical measures associated with this protocol are determined by how $\E$ would disturb a reference equilibrium state $\tau$ to some different state $\tau'$, if the protocol had been applied to this reference state instead of the $n$-copy magic state.

@@

The above bound depends on: (a) quantum computational measures $\epsilon, \epsilon'$, (b) thermodynamic quantities $F, F',$  and (c) intermediate terms $\phi, \phi'$. The intermediate terms depend on how the energy eigenbasis of the system $\{|E_k\>\}$ relates to the computational stabilizer basis $\{|k\>\}$.  In particular, the energetic term $\phi$ quantifies the degree to which the negativity in the Wigner representation can be ascribed a sharp energetic value for the given Hamiltonian of the system. Its form is similar to $F$, and so can informally be viewed as a `magic free energy' term, associated to a `magic partition function' $\zeta$. The coefficients $\alpha_k$ may be negative for some $k$, when the Hamiltonian eigen-decomposes into non-stabilizer states, but the quantity $\phi$ is always well-defined, since the function $\zeta$ is a positive multiple of the stabilizer Wigner component $W_\tau(0,0)$, as seen in the proof of the theorem.

The sole purpose of the Clifford unitaries $C_1,C_2$ is to simplify the form of the $\alpha_k$ coefficients. \nick{That's not right, without Clifford processing we would get bounds that depend on $u$ whenever $\x_\star$ corresponds to $u$, as well as $v$ whenever $\x_\star$ corresponds to $v$ as seen in the contour plot - these are different values} For example, if $H=Z$ then the energy eigenbasis is equivalent to the computational basis up to a Fourier transform, and so $\phi$ would have a sharp, temperature-independent value $\phi \in \{E_0,E_1,E_2\}$ once we account for the Clifford change in basis.

@@

The only assumption that cannot be trivially dropped is the assumption that we can ignore correlations in the thermodynamic limit. This is a common situation in thermodynamics, where correlations are often sub-linear in the particle number in the macroscopic limit. One could perturb about this scenario, and make use of variational tools such as the Bogoliubov inequality~\cite{bogolyubov_1966} for approximating the free energy of a system, to obtain bounds in a similar form.

@@

\null\newpage
\null\newpage
\section{Outlook and open questions}
\label{sec:lower_bounds}

\ddd{[Initial text. To be edited.]}
We have described how relative majorization can be used to establish upper bounds on magic distillation protocols that take into account additional physics of the system. Our bounds only exploited the very simplest aspects of the Lorenz curves of the quasi-distributions, so it would be of interest to sharpen these bounds and exploit more of the structure in the asymptotic limit. This raises interesting questions that do no arise for probability distributions, because we no longer have a notion of typicality and the central limit theorem does not apply. That said, for special states such as the Strange state the asymptotic behaviour is relatively simple, so one could study this and obtain a better handle on the asymptotic behaviour ($n \rightarrow \infty$). An example of such a question was raised regarding the numerical bounds plotted in~\cref{fig:distill_bounds} for the unital fragment. In~\cref{app:lcsu_technical} we give some more extensive analysis in the unital fragment that could be explored further, while in~\cref{app:lcst_technical} we also provide some tools which are useful in all fragments.

@@

Here, we present basic properties of the phase-point operators and the Wigner distribution that are used throughout the paper.

@@

\section{Properties of majorization}
\label{app:major}

We now give the following equivalent formulations of relative majorization.

@@

We now prove the following result, which was stated in~\cref{sec:monotones_frag}.

@@

\section{Technical properties of Lorenz curves and $\sigma$--fragments}
\label{app:frag}

\subsection{First elbow constraint}
\label{app:elb_constraint}
Here we prove a simple majorization constraint, arising by considering only the ascending part of the Lorenz curves between the origin $(0,0)$ and the first elbow.
\begin{proposition}\label{prop:first_elb}
	Consider a magic state process $\rho \longrightarrow \rho'$ with input and output Lorenz curves $\lc{\rho}{\sigma}(x), \lc{\rho'}{\sigma'}(x)$ and denote the coordinates of the first elbow of $L_{\rho|\sigma}(x)$ by $(X_0, L_0)$ and the coordinates of the first elbow of $L_{\rho' |\sigma'}(x)$ by $(X'_0, L'_0)$.

@@

\subsection{Lorenz curve constraint independence}

We now prove a result on the independence of the Lorenz curve constraints, stated in~\cref{sec:major_frag}.
\begin{proposition}\label{thm:elbows}
	Let $\rho, \rho'$ be two quantum states with Lorenz curves $\lc{\rho}{\sigma}(x), \lc{\rho'}{\sigma'}(x)$ in the $\sigma$--fragment.

@@

\subsection{Technical properties of $\sigma$--fragments}

\begin{proposition}\label{thm:frag_app}
    Let $\R = (\F, \O)$ be a magic theory and $\sigma, \sigma' \in \F$. The following statements hold:
    \begin{enumerate}
        \item No $\sigma$--fragment of $\R$ is empty.
        \item If a free operation leaves two states invariant, then it also leaves their mixtures invariant, 
        \begin{equation*}
            \O_{\sigma} \cap \O_{\sigma'} \subseteq \O_{p\sigma + (1-p)\sigma'}\ \text{for any}\ p \in [0,1].
        \end{equation*}
    \end{enumerate}
\end{proposition}
\begin{proof}$ $\vspace{-12pt}\\

\begin{enumerate}
    \item The identity channel $1_{\rm{C}}$ belongs to every $\sigma$--fragment, as $1_{\rm{C}} \in \O$ and $1_{\rm{C}}\sigma = \sigma$ for all $\sigma \in \F$.

    \item Let $\E \in \O_{\sigma} \cap \O_{\sigma'}$.
    Then $\E \in \cptp$ and corresponds to stochastic Wigner distribution $\W{\E}$ such that $\W{\E} \W{\sigma} = \W{\sigma}$ and $\W{\E} \W{\sigma'} = \W{\sigma'}$.
    Then, $\W{\E} \W{p\sigma + (1-p)\sigma'} = \W{p\sigma + (1-p)\sigma'}$ for any $p \in [0,1]$ due to the additive property~\ref{en:w4} of the Wigner distribution, implying that state $p\sigma + (1-p)\sigma'$ is also left invariant by $\E$.
\end{enumerate}
\vspace{-20pt}
\end{proof}

@@

\section{Lorenz curves in the unital fragment}
\label{app:lcsu_technical}

@@

\newpage
We can get explicit expressions for all $9^{n}$ points of the Lorenz curve $\lc{\rho_{\rm{S}}(\epsilon)^{\otimes n}}{(\id/3)^{\otimes n}}$, in terms of the elbow coordinates:
\begin{align}
    x_{ij} &= \left( 1-\frac{j}{m_{i}} \right) x_{i-1} + \frac{j}{m_{i}} x_{i}, \label{eq:x}\\
    L_{ij} &= \left( 1-\frac{j}{m_{i}} \right) L_{i-1} + \frac{j}{m_{i}} L_{i} \label{eq:l}
\end{align}
for $j = 1,\dots,m_{i}$ and $i=0,\dots,n$, where multiplicities $m_i$ are given in~\cref{tab:lcsu}.

@@

The following result shows that if the energy eigenbasis is a stabilizer basis then things simplify more.
\begin{proposition} 
	Let $H$ be a Hamiltonian $H$ with spectrum $\{E_k\}_{k \in \mathbb{Z}_d}$.
	If $H$ has a stabilizer eigenbasis, then $\phi = E_{k_{\star}}$ for some $k_{\star} \in \mathbb{Z}_d$.
\end{proposition}
\begin{proof}
	We first write $\alpha_k$ as
\begin{equation}
	\alpha_k = \<E_k | A_{\bmx_{\star}} |E_k\> = d W_{\ketbra{E_k}}(\bmx_{\star})
\end{equation}
Since each $\ketbra{E_k}$ is a stabilizer state, their Wigner distributions have orthogonal supports in the phase space, so only one of them, say $k = k_{\star}$ contains point $\bmx_{\star}$ in its support. 
Additionally, its distribution is uniform on its support, so $W_{\ketbra{E_{k_{\star}}}}(\bmx_{\star}) = 1/d$.

>>>>

\nick{re-write based on new relative majorisation approach}
The past few years have seen rapid progress towards the goal of a fault-tolerant quantum computer~\cite{Fowler_2012, Herrera_2010, Nickerson_2014, Nikahd_2017, chao_2018, lin_pieceable_2020, Lin_2020, Bourassa_2021} \ddd{[delete the first two from the bibliography - they are ancient and the second is mine, so it looks like bad self-citing]}. However, many challenges remain and there is increasing need for theory to take into account physical limitations of the hardware involved. The surface code~\cite{Bravyi_1998, Freedman_2001, Dennis_2002, Raussendorf_2007} is a leading framework for fault-tolerance with very high error thresholds. Within this scheme, Clifford unitaries can be implemented in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem~\cite{Eastin_2009}, we also know that it is impossible to have a universal set of transversal gates, and although Clifford unitaries can be realised transversally~\cite{Calderbank_1996, Steane_1996}, one needs to find ways around the Eastin-Knill restriction. This can be achieved by injecting in quantum states, called magic states, which promote the Clifford group to universal quantum computing~\cite{cit:bravyi}. The obstacle to this is that these states are invariably noisy and so protocols involving stabilizer operations must be employed to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question then arises about the overhead on purifying many copies of a magic state into less noisy forms.

@@

We approach this problem using tools from statistical mechanics and recent work from the overlap of thermodynamics and quantum information theory~\cite{cit:gour, cit:gour2}\nick{You had a specific review in mind I think, which one?} \ddd{[cite the Danzig paper from the early 2000s cite thermo-majorization Horodecki Oppenheim, cite the ``second laws'' paper by Brandao et al, cite Matteo's review, cite review articles by Goold et al, and Janet Anders]}. In particular, we shall view \emph{magic as a form of free energy}, relative to the set of stabilizer states. This perspective takes the convex set of stabilizer states as the set of ``equilibrium'' states, and magic corresponds to a form of non-equilibrium free energy that is strictly non-classical.

@@

\begin{figure}[t]
    \centering
        \includegraphics[scale=0.4]{figs/protocol_diagram.pdf}
    \caption{\textbf{Analysing magic protocols relative to a stabilizer reference state.} 
	A magic protocol aims to turn $n$ copies of some magic state $\rho_1$ to $m$ copies of some other state $\rho_2$.
	Constraints on the efficiency of the protocol can be imposed by considering its action on an separate probe state $\tau$.
	\nick{How do we improve the illustration?}\ddd{[I would have two separate box transformations, the top one has $\rho$ fed into $\E$, the bottom one has $\tau$ fed into it. Then put the relevant property (e.g. $\epsilon$ for $\rho$, and $F$ for $\tau$ next to the states. If viable then have $\phi$ somehow linking the top and bottom -- conveying it is a hybrid quantity in some sense.]}
    }
    \label{fig:sketch}
\end{figure}

@@

We now consider a magic distillation protocol on multiple identical qudits in a noisy magic state $\rho(\epsilon)$, with noise parameter $\epsilon$, sending 
\begin{equation}
\rho(\epsilon)^{\otimes n} \longrightarrow \E(\rho(\epsilon)^{\otimes n}) =\rho(\epsilon')^{\otimes m}
\end{equation}
with $n \gg 1$ \nick{Should we replace $m$ with $R n$ or $n'$ throughout?}\ddd{[Nah, $m$ is simpler than $Rn$ -- you gotta resist the urge to complicate expressions, unless it's needed. *Bad* habit.]}, $\epsilon' <\epsilon$ and $\E$ denoting the quantum channel induced by the protocol. We can now provide distillation bounds that depend on energetic and entropic aspects of the protocol. We do this relative to a reference temperature that can be chosen freely. The key idea is that the physical measures associated with this protocol are determined by how $\E$ would disturb a reference equilibrium state $\tau$ to some different state $\tau'$, if the protocol had been applied to this reference state instead of the $n$-copy magic state.

@@

The above bound depends on: (a) quantum computational measures $\epsilon, \epsilon'$, (b) thermodynamic quantities $F, F',$  and (c) intermediate terms $\phi, \phi'$. The intermediate terms depend on how the energy eigenbasis of the system $\{|E_k\>\}$ relates to the computational stabilizer basis $\{|k\>\}$.  In particular, the energetic term $\phi$ quantifies the degree to which the negativity in the Wigner representation can be ascribed a sharp energetic value for the given Hamiltonian of the system. Its form is similar to $F$, and so can informally be viewed as a `magic free energy' term, associated to a `magic partition function' $\zeta$. The coefficients $\alpha_k$ may be negative for some $k$, when the Hamiltonian eigen-decomposes into non-stabilizer states, but the quantity $\phi$ is always well-defined, since the function $\zeta$ is a positive multiple of the stabilizer Wigner component $W_\tau(0,0)$, as seen in the proof of the theorem. In the case that the system's Hamiltonian $H$ has an eigenbasis consisting of stabilizer states then we show in Appendix \ref{sharp-phi} that $\phi$ takes on a sharp, temperature-independent value $\phi \in \{E_0,E_1,E_2\}$.

The primary role of the pre/post-processing by Clifford unitaries is to simplify the form of the bound, and allow it to be expressed in terms of free energy quantities that do not depend on the parameters $\epsilon, \epsilon', \beta$ in a complex form.

@@

The only assumption that is non-trivial is the assumption that we can ignore correlations in the thermodynamic limit. This is a common situation in thermodynamics, where correlations are often sub-linear in the particle number in the macroscopic limit. One could perturb about this scenario, and make use of variational tools such as the Bogoliubov inequality~\cite{bogolyubov_1966} for approximating the free energy of a system, to obtain bounds in a similar form.

@@

\section{Outlook and open questions}
\label{sec:lower_bounds}

We have described how relative majorization can be used to establish upper bounds on magic distillation protocols that take into account additional physics of the system. Our bounds only exploited the very simplest aspects of the Lorenz curves of the quasi-distributions, so it would be of interest to sharpen these bounds and exploit more of the structure in the asymptotic limit. This raises interesting questions that do no arise for probability distributions, because we no longer have a notion of typicality and the central limit theorem does not apply. That said, for special states such as the Strange state the asymptotic behaviour is relatively simple, so one could study this and obtain a better handle on the asymptotic behaviour ($n \rightarrow \infty$). An example of such a question was raised regarding the numerical bounds plotted in~\cref{fig:distill_bounds} for the unital fragment. In~\cref{app:lcsu_technical} we give some more extensive analysis in the unital fragment that could be explored further, while in~\cref{app:lcst_technical} we also provide some tools which are useful in all fragments.

@@

In this appendix we present basic properties of the phase-point operators and the Wigner distribution that are used throughout the paper.

@@

\section{Properties of relative majorization and Lorenz curves}
\label{app:major}

The following provides equivalent formulations of relative majorization.

@@

\section{Lorenz curves and $\sigma$--fragments}
\label{app:frag}

We now prove the following result on the relation between majorization in a particular fragment and mana, which was stated in~\cref{sec:monotones_frag}.

@@

We now prove a simple majorization constraint, arising by considering only the part of the Lorenz curves between the origin $(0,0)$ and the first elbow.
\begin{proposition}\label{prop:first_elb}
	Consider a magic state process $\rho \longrightarrow \rho'$ with input and output Lorenz curves $\lc{\rho}{\sigma}(x), \lc{\rho'}{\sigma'}(x)$ and denote the coordinates of the first elbow of $L_{\rho|\sigma}(x)$ by $(X_0, L_0)$ and the coordinates of the first elbow of $L_{\rho' |\sigma'}(x)$ by $(X'_0, L'_0)$.

@@

We now show that the relative majorization condition can be reduced to considering the elbows of the Lorenz curves.
\begin{proposition}\label{thm:elbows}
	Let $\rho, \rho'$ be two quantum states with Lorenz curves $\lc{\rho}{\sigma}(x), \lc{\rho'}{\sigma'}(x)$ in the $\sigma$--fragment.

@@

@@

\section{Lorenz curves for unital processes}
\label{app:lcsu_technical}

@@

We can get explicit expressions for all $9^{n}$ points of the Lorenz curve $\lc{\rho_{\rm{S}}(\epsilon)^{\otimes n}}{(\id/3)^{\otimes n}}$, in terms of the elbow coordinates:
\begin{align}
    x_{ij} &= \left( 1-\frac{j}{m_{i}} \right) x_{i-1} + \frac{j}{m_{i}} x_{i}, \label{eq:x}\\
    L_{ij} &= \left( 1-\frac{j}{m_{i}} \right) L_{i-1} + \frac{j}{m_{i}} L_{i} \label{eq:l}
\end{align}
for $j = 1,\dots,m_{i}$ and $i=0,\dots,n$, where multiplicities $m_i$ are given in~\cref{tab:lcsu}.

@@

The following result shows that if the energy eigenbasis is a stabilizer basis then things simplify more.
\begin{proposition} \label{sharp-phi}
	Let $H$ be a Hamiltonian $H$ with spectrum $\{E_k\}_{k \in \mathbb{Z}_d}$.
	If $H$ has a stabilizer eigenbasis, then $\phi = E_{k_{\star}}$ for some $k_{\star} \in \mathbb{Z}_d$.
\end{proposition}
\begin{proof}
	We first write $\alpha_k$ as
\begin{equation}
	\alpha_k = \<E_k | A_{\bmx_{\star}} |E_k\> = d W_{\ketbra{E_k}}(\bmx_{\star})
\end{equation}
Since each $\ketbra{E_k}$ is a stabilizer state, their Wigner distributions have orthogonal supports in the phase space, so only one of them, say $k = k_{\star}$ contains point $\bmx_{\star}$ in its support. 
Additionally, its distribution is uniform on its support, so $W_{\ketbra{E_{k_{\star}}}}(\bmx_{\star}) = 1/d$.
